---
title: "a3_task3_mia_guarnieri"
author: "Mia Guarnieri"
date: "10/29/2021"
output: 
  html_document: 
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Task 3

## Overview
This code wrangles, analyzes, and visualizes data about the size frequency distribution of red and purple sea urchins recorded by Santa Barbara Coastal Long Term Ecological Research program divers at five long-term kelp removal experiment sites. Shell diameter was measured in centimeters for 50 of each type of urchin.

**Data citation:** Reed, D. 2020. SBC LTER: Reef: Long-term experiment: Kelp removal: Urchin size frequency distribution ver 20. Environmental Data Initiative. https://doi.org/10.6073/pasta/5a1e9ef03aa47bd2225c0bb98a02a63b. 

## Setup

```{r}
#attach packages
library(tidyverse)
library(here)
library(janitor)
library(kableExtra)

#read in the data
urchins <- read.csv(here("Data", "urchins.csv"))

#calling up the dtaframe for reference
head(urchins)
```


## Size comparisons of red and purple urchins

### Histogram

This faceted histogram compares the distribution of sizes between purple and red urchins across both treatments. 

```{r}
ggplot(data = urchins, aes(x = SIZE)) + #designating the data and x variable
  geom_histogram(bins = 10, color = "gray0", aes(fill = COMMON_NAME)) + #selecting a histogram, designating bin number, color coding by species, and adding outlines
  scale_fill_manual(values = c("darkviolet", "firebrick3")) + #designating color coding
  facet_wrap(~COMMON_NAME) + #splitting the graph by common name
  labs(x = "Size (cm)", #labeling axes
       y = "Count")
```

**Figure 1**: A histogram of purple and red sea urchin sizes, measured in centimeters.

### Data table of urchin statistics

The following data table presents summary statistics for red and purple urchin samples, including mean, median, standard deviation, and sample size.

```{r}

```











